---
title: "mixed effects"
author: "Mingjie"
date: "7/4/2021"
output: html_document
---

```{r setup, eval=TRUE, include=FALSE}
# boilerplate package loading chunk

# script-specific packages
suppressPackageStartupMessages({
  library(sjPlot)
  library(broom.mixed)
  library(lme4)
  library(MuMIn)
  library(lmtest)
  library(lattice)
  library(equatiomatic)
  library(DataExplorer)
  library(mvabund)
})

# general-use packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(fs)
  library(glue)
  library(readxl)
  library(cowplot)
  library(lubridate)
  library(patchwork)
  library(broom)
  library(ggeffects)
  library(viridis)
  library(arsenal) # for tableby()
  library(waldo) # for compare()
  library(sjmisc) # for rotate_df()
  library(envDocument)
  library(inspectdf)
  library(conflicted)
  library(knitr)
  library(beepr)
  library(pivottabler)
  library(furrr)
  library(scales)
  library(janitor)
  library(tictoc)
  library(ragg)
})

# Sometimes, two or more packages use the same function names. The {conflicted} package lets you set which package gets precedence. For example, the next line enforces that filter() refers to the {dplyr} package. If you want to use the command filter() from a different package, you just need to precede it with the desired package name like this: stats::filter.
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("mutate", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("summarise", "dplyr", quiet = TRUE)
conflict_prefer("first", "dplyr", quiet = TRUE)
conflict_prefer("here", "here", quiet = TRUE)
conflict_prefer("separate", "tidyr", quiet = TRUE)
conflict_prefer("unite", "tidyr", quiet = TRUE)
conflict_prefer("trim", "sjmisc", quiet = TRUE)
conflict_prefer("rescale", "scales", quiet = TRUE)
conflict_prefer("intersect", "dplyr", quiet = TRUE)
conflict_prefer("setdiff", "dplyr", quiet = TRUE) # w/out this, R crashes
conflict_prefer("to_factor", "sjmisc", quiet = TRUE)
conflict_prefer("trim", "glue", quiet = TRUE)
conflict_prefer("discard", "purrr", quiet = TRUE)
conflict_prefer("extract", "tidyr", quiet = TRUE)
conflict_prefer("col_factor", "readr", quiet = TRUE)

# R version
R.version.string
# Print real numbers, not scientific notation.
options(scipen = 999)
```

```{r load and explore data}
here()
mydata <- read_table(here("data", "mocksoup_longtable.txt"))

# introduce(mydata)
# plot_intro(mydata)
# plot_missing(mydata)
# plot_bar(mydata)
# plot_histogram(mydata)
```

```{r log predictors and responses}
mydata$log.readNum <- log(mydata$readNum + 1)
mydata$log.inputCOINum <- log(mydata$inputCOINum + 1)
mydata$log.mtagNum <- log(mydata$mtagNum + 1)
mydata$log.inputGDNA <- log(mydata$inputGDNA + 1)
mydata$log.mtagNumSpikeCorr <- log(mydata$mtagNumSpikeCorr + 1)
mydata$log.readNumSpikeCorr <- log(mydata$readNumSpikeCorr + 1)
plot_missing(mydata) # some PCRs produced 0 reads, set to zero with log(x+1)
plot_histogram(mydata) # these OTUs don't have inputCOINum values
str(mydata)

# not run
# mydata <- mydata %>%
#   #   filter(readNum != 0) %>% 
#   filter(!OTUID %in% c("HAP222", "HAP264")) 
```

################################ mixed effects model #############################

```{r COI spike-corrected readNum}
# test random effects
lmer_coi_readSpikeCorr1 <- lmer(log.inputCOINum ~ log.readNumSpikeCorr +
  (log.readNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = FALSE)
summary(lmer_coi_readSpikeCorr1)
plot(lmer_coi_readSpikeCorr1)

lmer_coi_readSpikeCorr2 <- lmer(log.inputCOINum ~ log.readNumSpikeCorr +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = FALSE)
summary(lmer_coi_readSpikeCorr2)
plot(lmer_coi_readSpikeCorr2)

lmer_coi_readSpikeCorr3 <- lmer(log.inputCOINum ~ log.readNumSpikeCorr +
   (1 | soupRep/pcrRep), data = mydata, REML = FALSE)
# singular fit
summary(lmer_coi_readSpikeCorr3)
plot(lmer_coi_readSpikeCorr3)

AIC(lmer_coi_readSpikeCorr1, lmer_coi_readSpikeCorr2, lmer_coi_readSpikeCorr3) # lmer_coi_readSpikeCorr1 has lower AIC, keep random intercept and slope, keep OTUID, soupRep, pcrRep

# test fixed effects
lmer_coi_readSpikeCorr4 <- lmer(log.inputCOINum ~ 1 +
  (log.readNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)

AIC(lmer_coi_readSpikeCorr1, lmer_coi_readSpikeCorr4) # lmer_coi_readSpikeCorr1
anova(lmer_coi_readSpikeCorr1, lmer_coi_readSpikeCorr4) # significantly different. keep fixed effect

# parameter estimates
lmer_coi_readSpikeCorr5 <- lmer(log.inputCOINum ~ log.readNumSpikeCorr +
  (log.readNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = TRUE)
summary(lmer_coi_readSpikeCorr5)
r.squaredGLMM(lmer_coi_readSpikeCorr5) # 0.416 0.96

sresid <- resid(lmer_coi_readSpikeCorr5, type = "pearson")
hist(sresid)
plot(lmer_coi_readSpikeCorr5)

pred.mm1 <- ggpredict(lmer_coi_readSpikeCorr5, terms = c("log.readNumSpikeCorr"))

(p1 <- ggplot() +
  geom_line(
    data = pred.mm1, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) +
  geom_ribbon(
    data = pred.mm1, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.3
  ) +
  geom_point(data = mydata, aes(
    x = log.readNumSpikeCorr, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.readNumSpikeCorr, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(x = "ln(spike-corrected OTU size)", y = "ln(input COI copy number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r COI non-spike-corrected readNum}
lmer_coi_read1 <- lmer(log.inputCOINum ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_coi_read2 <- lmer(log.inputCOINum ~ log.readNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_coi_read3 <- lmer(log.inputCOINum ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
AIC(lmer_coi_read1, lmer_coi_read2, lmer_coi_read3) # choose lmer_coi_read1

# test fixed effect
lmer_coi_read4 <- lmer(log.inputCOINum ~ 1 +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
AIC(lmer_coi_read1, lmer_coi_read4)
anova(lmer_coi_read1, lmer_coi_read4) # keep fixed effect

# estimate parameters with lmer_coi_read1
lmer_coi_read5 <- lmer(log.inputCOINum ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = TRUE)
# singular fit
summary(lmer_coi_read5)
r.squaredGLMM(lmer_coi_read5) # MuMIn package 0.042 0.85
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_coi_read5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)

plot(lmer_coi_read5) # gives a heteroscedasticity plot

pred.mm2 <- ggpredict(lmer_coi_read5, terms = c("log.readNum"))

# Plot the predictions
(p2 <- ggplot() +
  geom_line(
    data = pred.mm2, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + # slope
  geom_ribbon(
    data = pred.mm2, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.readNum, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.readNum, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(x = "ln(non-spike-corrected OTU size)", y = "ln(input COI copy number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r gDNA spike-corrected readNum}
lmer_gDNA_readSpikeCorr1 <- lmer(log.inputGDNA ~ log.readNumSpikeCorr +
   (log.readNumSpikeCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata, REML = FALSE)

lmer_gDNA_readSpikeCorr2 <- lmer(log.inputGDNA ~ log.readNumSpikeCorr +
   (1 | OTUID) +
   (1 | soupRep / pcrRep), data = mydata, REML = FALSE)

lmer_gDNA_readSpikeCorr3 <- lmer(log.inputGDNA ~ log.readNumSpikeCorr +
   (1 | soupRep / pcrRep), data = mydata, REML = FALSE)
# singular fit

AIC(lmer_gDNA_readSpikeCorr1, lmer_gDNA_readSpikeCorr2, lmer_gDNA_readSpikeCorr3) # lmer_gDNA_readSpikeCorr1 has lower AIC, use random intercept and slope, keep OTUID, soupRep, pcrRep

# test fixed effects
lmer_gDNA_readSpikeCorr4 <- lmer(log.inputGDNA ~ 1 +
   (log.readNumSpikeCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata, REML = F)

AIC(lmer_gDNA_readSpikeCorr1, lmer_gDNA_readSpikeCorr4) # choose lmer_gDNA_readSpikeCorr1
anova(lmer_gDNA_readSpikeCorr1, lmer_gDNA_readSpikeCorr4) # significantly different. keep fixed effect


# fit model
lmer_gDNA_readSpikeCorr5 <- lmer(log.inputGDNA ~ log.readNumSpikeCorr +
   (log.readNumSpikeCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata, REML = TRUE)
summary(lmer_gDNA_readSpikeCorr5)
r.squaredGLMM(lmer_gDNA_readSpikeCorr5) # 0.531 0.96
#
sresid <- resid(lmer_gDNA_readSpikeCorr5, type = "pearson")
hist(sresid)
plot(lmer_gDNA_readSpikeCorr5)
plot(sresid ~ mydata$log.readNumSpikeCorr)
#
pred.mm3 <- ggpredict(lmer_gDNA_readSpikeCorr5, terms = c("log.readNumSpikeCorr"))
#
(p3 <- ggplot() +
  geom_line(
    data = pred.mm3, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + 
  geom_ribbon(
    data = pred.mm3, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.readNumSpikeCorr, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.readNumSpikeCorr, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  ylim(2.3, 4.7) +
  labs(x = "ln(spike-corrected OTU size)", y = "ln(input genomic DNA)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r gDNA non-spike-corrected readNum}
lmer_gDNA_read1 <- lmer(log.inputGDNA ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_gDNA_read2 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_gDNA_read3 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit

AIC(lmer_gDNA_read1, lmer_gDNA_read2, lmer_gDNA_read3) # lmer_gDNA_read3

# fixed effect
lmer_gDNA_read4 <- lmer(log.inputGDNA ~ 1 +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
AIC(lmer_gDNA_read3, lmer_gDNA_read4) # keep fixed effect
anova(lmer_gDNA_read3, lmer_gDNA_read4) 


######## final model###########
lmer_gDNA_read5 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata, REML = T)
# singular fit
r.squaredGLMM(lmer_gDNA_read5) # 0.005 0.005
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_gDNA_read5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)

plot(lmer_gDNA_read5) # gives a heteroscedasticity plot

plot(sresid ~ mydata$log.readNum)

pred.mm4 <- ggpredict(lmer_gDNA_read5, terms = c("log.readNum"))

# Plot the predictions
(p4 <- ggplot() +
  geom_line(
    data = pred.mm4, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + # slope
  geom_ribbon(
    data = pred.mm4, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.readNum, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.readNum, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  ylim(2.3, 4.7) +
  labs(x = "ln(non-spike-corrected OTU size)", y = "ln(Input Genomic DNA)") +
  theme_cowplot() +
  theme(legend.position = "none"))

(p4 + p3)
(p2 + p1) / (p4 + p3)
```


```{r gDNA non-spike-corrected UMI number}
lmer_gDNA_UMI1 <- lmer(log.inputGDNA ~ log.mtagNum +
  (log.mtagNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_gDNA_UMI2 <- lmer(log.inputGDNA ~ log.mtagNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_gDNA_UMI3 <- lmer(log.inputGDNA ~ log.mtagNum +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit

AIC(lmer_gDNA_UMI1, lmer_gDNA_UMI2, lmer_gDNA_UMI3) # lmer_gDNA_UMI1

# fixed effect
lmer_gDNA_UMI4 <- lmer(log.inputGDNA ~ 1 +
  (log.mtagNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
AIC(lmer_gDNA_UMI1, lmer_gDNA_UMI4) # lmer_gDNA_UMI1
anova(lmer_gDNA_UMI1, lmer_gDNA_UMI4) # lmer_gDNA_UMI1

lmer_gDNA_UMI5 <- lmer(log.inputGDNA ~ log.mtagNum +
  (log.mtagNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = T)

r.squaredGLMM(lmer_gDNA_UMI5) # MuMIn package 0.015 0.033
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_gDNA_UMI5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)

plot(lmer_gDNA_UMI5) # gives a heteroscedasticity plot

plot(sresid ~ mydata$log.mtagNum)

pred.mm5 <- ggpredict(lmer_gDNA_UMI5, terms = c("log.mtagNum"))

# Plot the predictions
(p5 <- ggplot() +
  geom_line(data = pred.mm5, aes(x = x, y = predicted),
            colour = "black", size = 1) + # slope
  geom_ribbon(
    data = pred.mm5, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.mtagNum, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.mtagNum, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(y = "ln(Input Genomic DNA)", x = "ln(non-spike-corrected UMI number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r gDNA spike-corrected UMI number}
lmer_gDNA_UMIspikeCorr1 <- lmer(log.inputGDNA ~ log.mtagNumSpikeCorr +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
lmer_gDNA_UMIspikeCorr2 <- lmer(log.inputGDNA ~ log.mtagNumSpikeCorr +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
lmer_gDNA_UMIspikeCorr3 <- lmer(log.inputGDNA ~ log.mtagNumSpikeCorr +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit

AIC(lmer_gDNA_UMIspikeCorr1, lmer_gDNA_UMIspikeCorr2, lmer_gDNA_UMIspikeCorr3) # lmer_gDNA_UMIspikeCorr1

# fixed effect
lmer_gDNA_UMIspikeCorr4 <- lmer(log.inputGDNA ~ 1 +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# failure to converge
AIC(lmer_gDNA_UMIspikeCorr1, lmer_gDNA_UMIspikeCorr4) # lmer_gDNA_UMIspikeCorr1

# parameter estimate
lmer_gDNA_UMIspikeCorr5 <- lmer(log.inputGDNA ~ log.mtagNumSpikeCorr +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = T)

r.squaredGLMM(lmer_gDNA_UMIspikeCorr5) # 0.52 0.94
sresid <- resid(lmer_gDNA_UMIspikeCorr5, type = "pearson")
hist(sresid)
plot(lmer_gDNA_UMIspikeCorr5)
plot(sresid ~ mydata$log.mtagNumSpikeCorr)

pred.mm6 <- ggpredict(lmer_gDNA_UMIspikeCorr5, terms = c("log.mtagNumSpikeCorr"))

(p6 <- ggplot() +
  geom_line(data = pred.mm6, aes(x = x, y = predicted),
            colour = "black", size = 1) + # slope
  geom_ribbon(
    data = pred.mm6, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.mtagNumSpikeCorr, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.mtagNumSpikeCorr, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  ylim(2.3, 4.7) + 
  labs(y = "ln(Input Genomic DNA)", x = "ln(spike-corrected UMI number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r COI non-spike-corrected UMI number}
lmer_COI_UMIspike1 <- lmer(log.inputCOINum ~ log.mtagNum +
  (log.mtagNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_COI_UMIspike2 <- lmer(log.inputCOINum ~ log.mtagNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
lmer_COI_UMIspike3 <- lmer(log.inputCOINum ~ log.mtagNum +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit

AIC(lmer_COI_UMIspike1, lmer_COI_UMIspike2, lmer_COI_UMIspike3) # lmer_gDNA_UMIspikeCorr1

# fixed effect
lmer_COI_UMIspike4 <- lmer(log.inputCOINum ~ 1 +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit
AIC(lmer_COI_UMIspike1, lmer_COI_UMIspike4) # lmer_COI_UMIspikeCorr1
anova(lmer_COI_UMIspike1, lmer_COI_UMIspike4) 

# parameter estimate
lmer_COI_UMIspike5 <- lmer(log.inputCOINum ~ log.mtagNum +
  (log.mtagNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = T)
# singular fit
r.squaredGLMM(lmer_COI_UMIspike5) # 0.054 0.85
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_COI_UMIspike5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)

plot(lmer_COI_UMIspike5) # gives a heteroscedasticity plot

# plot(sresid ~ mydata$log.mtagNum)

pred.mm7 <- ggpredict(lmer_COI_UMIspike5, terms = c("log.mtagNum"))

# Plot the predictions
(p7 <- ggplot() +
  geom_line(data = pred.mm7, aes(x = x, y = predicted),
            colour = "black", size = 1) + # slope
  geom_ribbon(
    data = pred.mm7, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.mtagNum, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.readNum, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(y = "ln(Input COI copy number)", x = "ln(non-spike-corrected UMI number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

```{r COI spike-corrected UMI number}
lmer_COI_UMIspikeCorr1 <- lmer(log.inputCOINum ~ log.mtagNumSpikeCorr +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
lmer_COI_UMIspikeCorr2 <- lmer(log.inputCOINum ~ log.mtagNumSpikeCorr +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
lmer_COI_UMIspikeCorr3 <- lmer(log.inputCOINum ~ log.mtagNumSpikeCorr +
  (1 | soupRep / pcrRep), data = mydata, REML = F)
# singular fit

AIC(lmer_COI_UMIspikeCorr1, lmer_COI_UMIspikeCorr2, lmer_COI_UMIspikeCorr3) # lmer_gDNA_UMIspikeCorr1

# fixed effect
lmer_COI_UMIspikeCorr4 <- lmer(log.inputCOINum ~ 1 +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = F)

AIC(lmer_COI_UMIspikeCorr1, lmer_COI_UMIspikeCorr4) # lmer_COI_UMIspikeCorr1
anova(lmer_COI_UMIspikeCorr1, lmer_COI_UMIspikeCorr4) 

# estimate parameters
lmer_COI_UMIspikeCorr5 <- lmer(log.inputCOINum ~ log.mtagNumSpikeCorr +
  (log.mtagNumSpikeCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata, REML = T)
r.squaredGLMM(lmer_COI_UMIspikeCorr5) # 0.434 0.95

sresid <- resid(lmer_COI_UMIspikeCorr5, type = "pearson")
hist(sresid)
plot(lmer_COI_UMIspikeCorr5)

pred.mm8 <- ggpredict(lmer_COI_UMIspikeCorr5, terms = c("log.mtagNumSpikeCorr"))

(p8 <- ggplot() +
  geom_line(data = pred.mm8, aes(x = x, y = predicted),
            colour = "black", size = 1) + # slope
  geom_ribbon(
    data = pred.mm8, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata, aes(
    x = log.mtagNumSpikeCorr, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata, aes(
      x = log.mtagNumSpikeCorr, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(y = "ln(Input COI copy number)", x = "ln(spike-corrected UMI number)") +
  theme_cowplot() +
  theme(legend.position = "none"))

p7 + p8
(p7 + p8) / (p5 + p6) 
```

```{r figures}
(p2 + p1) / (p7 + p8) + plot_annotation(tag_levels = 'A') # Figure 7
(p4 + p3) / (p5 + p6) + plot_annotation(tag_levels = 'A') # Figure 8
```

#### Test of Warton spike-in estimator on full dataset, with OTUID, soupRep, and pcrRep   

Upshot 1:  Warton's offset estimator is worse than a physical spike-in but better than not having it. With the offset estimator, the fixed-effect R2 = 10.3% for predicting input COI copy number (vs. 4.2% without) and R2 = 18.1% for predicting input genomic DNA (vs. 0.5% without). This is with 63 samples (= 7 soups X 3 soup replicates X 3 PCR replicates per soup) and 52 OTUs  

Upshot 2:  When i reduce the dataset to 7 soups (averaging over the 3 soup replicates and 3 PCR replicates per soup) and 52 OTUs, Warton's method no longer works, presumably because of too little data. Averaging the physical spike-in still works (shown elsewhere by Mingjie's analysis, not here)  

reformat data
```{r reformat to mydata2}
mydata2 <- mydata %>% 
  select(-mtagNum, -exp, -lib, -sampleName, -PCR, -readSpike, -mtagSpike,
         -readNumSpikeCorr, -mtagNumSpikeCorr, -inputCOINum, -inputGDNA,
         -starts_with("log.")) %>% 
  pivot_wider(names_from = OTUID, values_from = readNum)

otu <- mydata2 %>% 
  select(starts_with("HAP"))

X <- mydata2 %>% 
  select(!starts_with("HAP"))
```

Warton's spike-in estimator: offset0
```{r calculate offset0 for mydata2}
otu <- mvabund(otu)
# fit0 <- manyglm(otu ~ 1, family = "negative.binomial") # if no predictors
fit0 <- manyglm(otu ~ soup, family = "negative.binomial", data = X) # data = the env covariates. otu is the otu table
# fit0 <- manyglm(otu ~ soup, family = binomial("cloglog")) # for p/a data
offset0 <- log(rowSums(otu)) - log(rowSums(fitted(fit0))) 
mydata2$offset0 <- offset0
mydata2 <- mydata2 %>% 
  mutate(
    offset0_scale = rescale(offset0, to = c(1, 10))
  ) %>% 
  relocate(offset0, offset0_scale, .after = pcrRep) %>% 
  pivot_longer(cols = starts_with("HAP"), 
               names_to = "OTUID", 
               values_to = "readNum") %>% 
  mutate(
    readNumOffsetCorr = readNum / offset0_scale
  )

mydata2 <- mydata %>% 
  select(OTUID, soup, soupRep, pcrRep, inputCOINum, inputGDNA) %>% 
  left_join(mydata2) %>% 
  mutate(
    log.inputCOINum = log(inputCOINum + 1),
    log.inputGDNA = log(inputGDNA + 1),
    log.readNum = log(readNum + 1),
    log.readNumOffsetCorr = log(readNumOffsetCorr + 1)
  )

names(mydata2)
```

 [1] "OTUID"                 "soup"                  "soupRep"              
 [4] "pcrRep"                "inputCOINum"           "inputGDNA"            
 [7] "offset0"               "offset0_scale"         "readNum"              
[10] "readNumOffsetCorr"     "log.inputCOINum"       "log.inputGDNA"        
[13] "log.readNum"           "log.readNumOffsetCorr"

prediction of input COI copy number
```{r COI offset-corrected readNum}
# test random effects
lmer_coi_readOffsetCorr1 <- lmer(log.inputCOINum ~ log.readNumOffsetCorr +
  (log.readNumOffsetCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = FALSE)
summary(lmer_coi_readOffsetCorr1)
plot(lmer_coi_readOffsetCorr1)
# singular fit caused by zero variance in pcrRep:soupRep and soupRep, but these random factors must be retained because they are part of the experimental design
lmer_coi_readOffsetCorr2 <- lmer(log.inputCOINum ~ log.readNumOffsetCorr +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = FALSE)
summary(lmer_coi_readOffsetCorr2)
plot(lmer_coi_readOffsetCorr2)
# singular fit
lmer_coi_readOffsetCorr3 <- lmer(log.inputCOINum ~ log.readNumOffsetCorr +
   (1 | soupRep/pcrRep), data = mydata2, REML = FALSE)
# singular fit
summary(lmer_coi_readOffsetCorr3)
plot(lmer_coi_readOffsetCorr3)

AIC(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr2, lmer_coi_readOffsetCorr3) # lmer_coi_readOffsetCorr1 has lowest AIC, keep random intercept and slope, keep OTUID, soupRep, pcrRep

# test fixed effects
lmer_coi_readOffsetCorr4 <- lmer(log.inputCOINum ~ 1 +
  (log.readNumOffsetCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
AIC(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr4) # lmer_coi_readOffsetCorr1
anova(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr4) # significantly different. keep fixed effect

# parameter estimates
lmer_coi_readOffsetCorr5 <- lmer(log.inputCOINum ~ log.readNumOffsetCorr +
  (log.readNumOffsetCorr | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = TRUE)
# singular fit
summary(lmer_coi_readOffsetCorr5)
r.squaredGLMM(lmer_coi_readOffsetCorr5) # 0.1033 0.8676

sresid <- resid(lmer_coi_readOffsetCorr5, type = "pearson")
hist(sresid)
plot(lmer_coi_readOffsetCorr5)

pred.mm1 <- ggpredict(lmer_coi_readOffsetCorr5, 
                      terms = c("log.readNumOffsetCorr"))

(p9 <- ggplot() +
  geom_line(
    data = pred.mm1, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) +
  geom_ribbon(
    data = pred.mm1, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.3
  ) +
  geom_point(data = mydata2, aes(
    x = log.readNumOffsetCorr, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata2, aes(
      x = log.readNumOffsetCorr, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(x = "ln(Offset-corrected OTU size)", y = "ln(input COI copy number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

should be exactly the same as "COI non-spike-corrected readNum"
```{r COI non-offset-corrected readNum}
lmer_coi_read1 <- lmer(log.inputCOINum ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
lmer_coi_read2 <- lmer(log.inputCOINum ~ log.readNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
lmer_coi_read3 <- lmer(log.inputCOINum ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
AIC(lmer_coi_read1, lmer_coi_read2, lmer_coi_read3) # choose lmer_coi_read1

# test fixed effect
lmer_coi_read4 <- lmer(log.inputCOINum ~ 1 +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
AIC(lmer_coi_read1, lmer_coi_read4)
anova(lmer_coi_read1, lmer_coi_read4) # keep fixed effect

# estimate parameters with lmer_coi_read1
lmer_coi_read5 <- lmer(log.inputCOINum ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = TRUE)
# singular fit
summary(lmer_coi_read5)
r.squaredGLMM(lmer_coi_read5) # 0.042 0.85
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_coi_read5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)
plot(lmer_coi_read5) # gives a heteroscedasticity plot

pred.mm2 <- ggpredict(lmer_coi_read5, terms = c("log.readNum"))

# Plot the predictions
(p10 <- ggplot() +
  geom_line(
    data = pred.mm2, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + # slope
  geom_ribbon(
    data = pred.mm2, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata2, aes(
    x = log.readNum, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata2, aes(
      x = log.readNum, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(x = "ln(non-Offset-corrected OTU size)", y = "ln(input COI copy number)") +
  theme_cowplot() +
  theme(legend.position = "none"))

p10 + p9
```

prediction of input genomic DNA
```{r gDNA offset-corrected readNum}
lmer_gDNA_readOffsetCorr1 <- lmer(log.inputGDNA ~ log.readNumOffsetCorr +
   (log.readNumOffsetCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata2, REML = FALSE)
# singular fit
summary(lmer_gDNA_readOffsetCorr1)
lmer_gDNA_readOffsetCorr2 <- lmer(log.inputGDNA ~ log.readNumOffsetCorr +
   (1 | OTUID) +
   (1 | soupRep / pcrRep), data = mydata2, REML = FALSE)
# singular fit
lmer_gDNA_readOffsetCorr3 <- lmer(log.inputGDNA ~ log.readNumOffsetCorr +
   (1 | soupRep / pcrRep), data = mydata2, REML = FALSE)
# singular fit

AIC(lmer_gDNA_readOffsetCorr1, lmer_gDNA_readOffsetCorr2, lmer_gDNA_readOffsetCorr3) # lmer_gDNA_readOffsetCorr1 has lower AIC, use random intercept and slope, keep OTUID, soupRep, pcrRep

# test fixed effects
lmer_gDNA_readOffsetCorr4 <- lmer(log.inputGDNA ~ 1 +
   (log.readNumOffsetCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
AIC(lmer_gDNA_readOffsetCorr1, lmer_gDNA_readOffsetCorr4) # choose lmer_gDNA_readOffsetCorr1
anova(lmer_gDNA_readOffsetCorr1, lmer_gDNA_readOffsetCorr4) # significantly different. keep fixed effect


# fit model
lmer_gDNA_readOffsetCorr5 <- lmer(log.inputGDNA ~ log.readNumOffsetCorr +
   (log.readNumOffsetCorr | OTUID) +
   (1 | soupRep / pcrRep), data = mydata2, REML = TRUE)
# singular fit
summary(lmer_gDNA_readOffsetCorr5)
r.squaredGLMM(lmer_gDNA_readOffsetCorr5) # 0.181 0.49
#
sresid <- resid(lmer_gDNA_readOffsetCorr5, type = "pearson")
hist(sresid)
plot(lmer_gDNA_readOffsetCorr5)
plot(sresid ~ mydata2$log.readNumOffsetCorr)
#
pred.mm3 <- ggpredict(lmer_gDNA_readOffsetCorr5, 
                      terms = c("log.readNumOffsetCorr"))

(p11 <- ggplot() +
  geom_line(
    data = pred.mm3, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + 
  geom_ribbon(
    data = pred.mm3, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata2, aes(
    x = log.readNumOffsetCorr, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata2, aes(
      x = log.readNumOffsetCorr, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  ylim(2.3, 4.7) +
  labs(x = "ln(Offset-corrected OTU size)", y = "ln(input genomic DNA)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```

should be exactly the same as "gDNA non-spike-corrected readNum"
```{r gDNA non-offset-corrected readNum}
lmer_gDNA_read1 <- lmer(log.inputGDNA ~ log.readNum +
  (log.readNum | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
summary(lmer_gDNA_read1)
lmer_gDNA_read2 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | OTUID) +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
lmer_gDNA_read3 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit

AIC(lmer_gDNA_read1, lmer_gDNA_read2, lmer_gDNA_read3) # lmer_gDNA_read3

# fixed effect
lmer_gDNA_read4 <- lmer(log.inputGDNA ~ 1 +
  (1 | soupRep / pcrRep), data = mydata2, REML = F)
# singular fit
AIC(lmer_gDNA_read3, lmer_gDNA_read4) # keep fixed effect
anova(lmer_gDNA_read3, lmer_gDNA_read4) 


######## final model###########
lmer_gDNA_read5 <- lmer(log.inputGDNA ~ log.readNum +
  (1 | soupRep / pcrRep), data = mydata2, REML = T)
# singular fit
r.squaredGLMM(lmer_gDNA_read5) # 0.00487 0.005
# R2m = marginal R-squared (variance explained by fixed effects only)
# R2c = conditional R-squared (variance explained by random + fixed effects)

sresid <- resid(lmer_gDNA_read5, type = "pearson") # Extract the standardised residuals, check underline assumption of normally distibuted
hist(sresid)
plot(lmer_gDNA_read5) # gives a heteroscedasticity plot
plot(sresid ~ mydata2$log.readNum)

pred.mm4 <- ggpredict(lmer_gDNA_read5, terms = c("log.readNum"))

# Plot the predictions
(p12 <- ggplot() +
  geom_line(
    data = pred.mm4, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) + # slope
  geom_ribbon(
    data = pred.mm4, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.5
  ) +
  geom_point(data = mydata2, aes(
    x = log.readNum, y = log.inputGDNA,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata2, aes(
      x = log.readNum, y = log.inputGDNA,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  ylim(2.3, 4.7) +
  labs(x = "ln(non-Offset-corrected OTU size)", y = "ln(Input Genomic DNA)") +
  theme_cowplot() +
  theme(legend.position = "none"))

p11 + p12
```

```{r}
(p10 + p9) / (p12 + p11)
```

##### Test of Warton spike-in estimator on summarised dataset, averaged over soupRep and pcrRep  

reformat data
```{r reformat to mydata3}
mydata3 <- mydata %>% 
  select(-mtagNum, -exp, -lib, -sampleName, -PCR, -readSpike, -mtagSpike,
         -readNumSpikeCorr, -mtagNumSpikeCorr, -inputCOINum, -inputGDNA,
         -starts_with("log.")) %>% 
  group_by(OTUID, soup) %>% 
  summarise(
    readNumMean = round(mean(readNum))
  ) %>% 
  pivot_wider(names_from = OTUID, values_from = readNumMean)

otu <- mydata3 %>% 
  select(starts_with("HAP"))

X <- mydata3 %>% 
  select(!starts_with("HAP"))
```

Warton's spike-in estimator: offset0
```{r calculate offset0 for mydata3}
otu <- mvabund(otu)
# fit0 <- manyglm(otu ~ 1, family = "negative.binomial") # if no predictors
fit0 <- manyglm(otu ~ soup, family = "negative.binomial", data = X) # data = the env covariates. otu is the otu table
# fit0 <- manyglm(otu ~ soup, family = binomial("cloglog")) # for p/a data
offset0 <- log(rowSums(otu)) - log(rowSums(fitted(fit0))) 
mydata3$offset0 <- offset0
mydata3 <- mydata3 %>% 
  mutate(
    offset0_scale = rescale(offset0, to = c(1, 10))
  ) %>% 
  pivot_longer(cols = starts_with("HAP"), 
               names_to = "OTUID", 
               values_to = "readNumMean") %>% 
  mutate(
    readNumMeanOffsetCorr = readNumMean / offset0_scale
  ) %>% 
  relocate(OTUID, soup)


mydata4 <- mydata %>% 
  select(OTUID, soup, inputCOINum, inputGDNA) %>% 
  group_by(OTUID, soup) %>% 
  summarise(
    inputCOINum = round(mean(inputCOINum)),
    inputGDNA = round(mean(inputGDNA))
  ) 
  

mydata3 <- mydata3 %>% 
  right_join(mydata4) %>% 
  mutate(
    log.inputCOINum = log(inputCOINum + 1),
    log.inputGDNA = log(inputGDNA + 1),
    log.readNumMean = log(readNumMean + 1),
    log.readNumMeanOffsetCorr = log(readNumMeanOffsetCorr + 1)
  )  

names(mydata3)
```

 [1] "OTUID"                     "soup"                      "offset0"        
 [4] "offset0_scale"             "readNumMean"               "readNumMeanOffsetCorr"
 [7] "inputCOINum"               "inputGDNA"                 "log.inputCOINum"
[10] "log.inputGDNA"             "log.readNumMean"        "log.readNumMeanOffsetCorr"

mixed-effect model
results in a negative slope, which is the opposite of the expected result. The reason for the negative slope is that soup F is given the smallest offset. Dividing by the offset results in the largest readNumOffsetCorr values, when F soups should have the third smallest. 
```{r COI offset-corrected readNumMean}
# test random effects
lmer_coi_readOffsetCorr1 <- lmer(log.inputCOINum ~ log.readNumMeanOffsetCorr +
  (log.readNumMeanOffsetCorr | OTUID), data = mydata3, REML = FALSE)
summary(lmer_coi_readOffsetCorr1)
par(mfrow = c(2,2))
plot(lmer_coi_readOffsetCorr1)
par(mfrow = c(1,1))

lmer_coi_readOffsetCorr2 <- lmer(log.inputCOINum ~ log.readNumMeanOffsetCorr +
  (1 | OTUID), data = mydata3, REML = FALSE)
summary(lmer_coi_readOffsetCorr2)
plot(lmer_coi_readOffsetCorr2)


AIC(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr2) # lmer_coi_readOffsetCorr1 has lowest AIC, keep random intercept and slope

# test fixed effects
lmer_coi_readOffsetCorr4 <- lmer(log.inputCOINum ~ 1 +
  (log.readNumMeanOffsetCorr | OTUID), data = mydata3, REML = F)

AIC(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr4) # lmer_coi_readOffsetCorr1
anova(lmer_coi_readOffsetCorr1, lmer_coi_readOffsetCorr4) # significantly different. keep fixed effect

# parameter estimates
lmer_coi_readOffsetCorr5 <- lmer(log.inputCOINum ~ log.readNumMeanOffsetCorr +
  (log.readNumMeanOffsetCorr | OTUID), data = mydata3, REML = TRUE)
summary(lmer_coi_readOffsetCorr5)
r.squaredGLMM(lmer_coi_readOffsetCorr5) # 0.114 0.931

sresid <- resid(lmer_coi_readOffsetCorr5, type = "pearson")
hist(sresid)
plot(lmer_coi_readOffsetCorr5)

pred.mm1 <- ggpredict(lmer_coi_readOffsetCorr5, 
                      terms = c("log.readNumMeanOffsetCorr"))

(p9 <- ggplot() +
  geom_line(
    data = pred.mm1, aes(x = x, y = predicted),
    colour = "black", size = 1
  ) +
  geom_ribbon(
    data = pred.mm1, aes(
      x = x, ymin = predicted - std.error,
      ymax = predicted + std.error
    ),
    fill = "lightgrey", alpha = 0.3
  ) +
  geom_point(data = mydata3, aes(
    x = log.readNumMeanOffsetCorr, y = log.inputCOINum,
    colour = OTUID
  ), size = 0.8) +
  geom_smooth(
    data = mydata3, aes(
      x = log.readNumMeanOffsetCorr, y = log.inputCOINum,
      group = OTUID
    ),
    method = "lm", se = FALSE, size = 0.3, colour = "black", alpha = 0.5,
    linetype = 1
  ) +
  labs(x = "ln(Offset-corrected OTU size)", y = "ln(input COI copy number)") +
  theme_cowplot() +
  theme(legend.position = "none"))
```
